---
title: "Ampvis guide"
author: "Mads Albertsen"
date: "Friday, January 23, 2014"
output: html_document
---

# Installation

## Install R, Rtools and Rstudio
Install the latest version of [R, Rtools](http://www.r-project.org/) and [Rstudio](http://www.rstudio.com/).

## Install ampvis
A few packages needs to be manually installed using bioconducter, before ampvis can be installed. Make sure you run RStudio as `Administrator` to enable installation of all packages! If you are asked to update packages, then press "a" for all.
```{r install_packages, eval=FALSE}
install.packages("devtools")
source("http://bioconductor.org/biocLite.R")
biocLite("Biostrings")
biocLite("DESeq2")
biocLite("phyloseq")
```

Ampvis is installed directly from github using the devtools package. 
```{r install_ampvis, eval=FALSE}
library("devtools")
install_github("MadsAlbertsen/ampvis")
```

# Load ampvis and import data

## Load ampvis
```{r load_packages, message=FALSE, warning=FALSE}
library(ampvis)
```

## Load data
Data generated with `workflow scripts v.4+` can be loaded using the `amp_load` function. It converts the data into a single phyloseq object. First the raw data is loaded into R.
```{r load_data, eval=FALSE}
otutable <- read.delim(file = "data/otutable.txt", sep = "\t", header = T, check.names = F, row.names = 1)
metadata <- read.delim(file = "data/metadata.txt", header = T, sep = "\t")
refseq <- readDNAStringSet("data/otus.fa", format = "fasta")
```

... and then convert them to a single phyloseq object using the `amp_load` function. It also includes an optional rarefying step up front (I.e. subsample all samples to equal depth).
```{r load_phyloseq, eval=F}
d <- amp_load(otutable = otutable, 
              metadata = metadata, 
              refseq = refseq, 
              rarefy = 10000)
```

## MiDAS data
The MiDAS data is now directly included in ampvis and can be loaded using the `data()` function. It loads the data in the object `d`.
```{r load_MiDAS}
data(MiDAS_1.16)
```

The data is a phyloseq object with 575 samples and also includes the OTU sequences.
```{r see_MiDAS}
d
```

# Analyse the data

## Subset and filtering
The principle behind ampvis is that you first subset the data to what you want to look at using `phyloseq` and then visualise it using `ampvis`. Samples can be subset based on any available metadata. See the [phyloseq guide](http://joey711.github.io/phyloseq/) for more examples.
```{r see_variables}
sample_variables(d)
```

Here we'll subset to samples from `Aalborg West` and `Aalborg East` using the `Plant` variable in the metadata and store the data in the object `ds`.
```{r subset_samples}
ds <- subset_samples(d, Plant %in% c("Aalborg West", "Aalborg East"))
ds
```

Before we analyse the data it is often convinient to remove low abundant species. This can be done using the `filter_taxa` function from phyloseq. Here we keep OTUs that have been seen more than 9 times (of 10000) in at least 1 sample.
```{r filter_taxa}
dsf <- filter_taxa(ds, function(x) max(x) >= 10, TRUE)
dsf
```

The data can also be limited to specific taxonomic groups. Here I've limited the analysis to the genus Tetrasphaera.
```{r subset_taxa}
dsft <- subset_taxa(dsf, Genus == "g__Tetrasphaera")
dsft       
```

### Chaining it all together
The above functions can be combined more elegantly in order to make fewer temporary files using the `then` operator `%>%`. It takes the output from a given function and feeds it to the next. It is often called piping or chaining.
```{r pipe}
dstf2 <- subset_samples(d, Plant %in% c("Aalborg West", "Aalborg East")) %>%
         filter_taxa(function(x) max(x) >= 10, TRUE) %>%
         subset_taxa(Genus == "g__Tetrasphaera")
```

## Basic ampvis functions

### Heatmap
Looking at phylum level (default) on Aalborg East (AAE) and West (AAW) from 2006 to 2013. Proteobacteria have been split into the respective classes. Note that samples can be grouped based on any metadata variable. All functions have a basic help file that can be acessed using e.g. `?amp_heatmap`. 
```{r heatmap1, fig.width=10, fig.height=8, fig.align='center'}
amp_heatmap(data = dsf, 
            group = c("Plant", "Year"), 
            tax.class = "p__Proteobacteria")
```

The 25 most abundant Genera in AAW and AAE. The `tax.aggregate` parameter controls which level the taxonomic information is shown on. The data is sorted based on average abundance across all groups of samples.
```{r heatmap2, fig.width=10, fig.height=8, fig.align='center'}
amp_heatmap(data = dsf, 
            group = c("Plant", "Year"), 
            tax.class = "p__Proteobacteria",
            tax.aggregate = "Genus",
            tax.show = 25)
```

The 25 most abundant OTUs in AAW and AAE. You can always add higher taxonomic ranks to the names using the `tax.add` parameter. Here we also label each OTU with genus and phylum name. In addition, we scale the color using log10 instead of the default square root. If you don't like the numbers, you can just remove them using the 
```{r heatmap3, fig.width=12, fig.height=8, fig.align='center'}
amp_heatmap(data = dsf, 
            group = c("Plant", "Year"), 
            tax.class = "p__Proteobacteria",
            tax.aggregate = "OTU",
            tax.add = c("Phylum", "Genus"),
            tax.show = 25,
            plot.colorscale = "log10",
            plot.na = T,
            plot.numbers = F) 
```

### Boxplots
The heatmaps always show the average of the samples. The function `amp_rabund` (short for Rank Abundance) can be used to generate boxplots. The syntax is similar to the heatmap.
```{r boxplot1, fig.width=8, fig.height=8, fig.align='center'}
amp_rabund(data = dsf,
           tax.aggregate = "Genus",
           tax.add = "Phylum")
```

Boxplots can also be grouped.
```{r boxplot2, fig.width=9, fig.height=4, fig.align='center'}
amp_rabund(data = dsf,
           tax.aggregate = "Genus",
           tax.add = "Phylum", 
           tax.show = 10,
           group = "Plant",
           tax.class = "p__Proteobacteria")
```

The `amp_rabund` can also be used to generate cumulative rank abundance curves using the `plot.type` variable. These can be used to compare the diversity between samples. I.e. how many OTUs constitute 75% of the sample.
```{r boxplot3, fig.width=6, fig.height=5, fig.align='center'}
amp_rabund(data = dsf,
           plot.type = "curve",
           tax.aggregate = "OTU",
           group = "Plant")
```

### PCA
The function `amp_ordinate` can be used to make simple PCA plots (and NMDS) of your data. Notice that the PCA uses the OTU counts directly and not e.g. bray.curtis similarities.
```{r PCA1A, fig.width=8, fig.height=6, fig.align='center'}
amp_ordinate(data = dsf, 
             plot.color = "Plant")
```

The `plot.group` variable can be used to make the groupings for easy to see.
```{r PCA1B, fig.width=8, fig.height=6, fig.align='center'}
amp_ordinate(data = dsf, 
             plot.color = "Plant", 
             plot.group = "chull")
```

There is a large number of ways to visualise the data. I'll just show a few basic functions. You can for example add the n most influencial species. 
```{r PCA2, fig.width=8, fig.height=6, fig.align='center'}
amp_ordinate(data = dsf, 
             plot.color = "Plant", 
             plot.group = "chull",
             plot.nspecies = 10)
```

Trace the development over time. We also add the plant name directly to the plot using the `plot.group.label`. It plots the name of the group in the centroid of all samples in the group.
```{r PCA3, fig.width=8, fig.height=6, fig.align='center'}
amp_ordinate(data = dsf, 
             plot.color = "Year", 
             trajectory = "Date", 
             trajectory.group = "Plant",
             plot.group.label = "Plant")
```

.. or even test for significance of environmental factors. Here we also use the `output = "complete"`  option to get the data behind the plot so we can see the significance of the tested variables. Note: envifit only test if there is a correlation to the PCA's you display, here PC1 and PC2. There might be correlations hidden in the other PCA's.
```{r PCA4}
res <- amp_ordinate(data = dsf, 
                    plot.color = "Plant", 
                    envfit.factor = c("Plant","Period"), 
                    output = "complete")           
```

You can access, plots data and modelling results in the `res` object. Here we take a look to see if the fitting of Plant and Peroid was significant based on a permutation test.
```{r PCA4_res}
res$eff.model
```


Finally you can constrain the PCA and look for relationships to specific variables. Here we can see that there seem to be a shift in both plants to more Dechloromonas from 2006 to 2013.
```{r PC5, fig.width=8, fig.height=6, fig.align='center'}
amp_ordinate(data = dsf, 
             plot.color = "Year", 
             constrain = "Year", 
             plot.nspecies = 10)
```

### Core plots
The `amp_core` function can be used to investigate the core community in your samples. The default plot is a `frequency plot` that groups the OTUs based on how many samples they are seen in. If `weight` is set to `TRUE` then each OTU is weighted by their abundance. 
```{r core1, fig.width=6, fig.height=5, fig.align='center', message=FALSE}
amp_core(data = dsf, plot.type = "frequency", weight = T)
```

..if not it simply display the number of OTUs in each group.
```{r core2, fig.width=6, fig.height=5, fig.align='center', message=FALSE}
amp_core(data = dsf, plot.type = "frequency", weight = F)
```

You have to be carefull with all functions as your initial filtering will affect your plots. In the beginning we removed low abundant OTUs. However, if we do not filter the data the plots look a little different. Hence, we use the `ds` subset instead of the filtered version `dsf`. The weighted version does not look that different. 
```{r core3, fig.width=6, fig.height=5, fig.align='center', message=FALSE}
amp_core(data = ds, plot.type = "frequency", weight = T)
```

..but there are major changes to the unweighted version (counts).
```{r core4, fig.width=6, fig.height=5, fig.align='center', message=FALSE}
amp_core(data = ds, plot.type = "frequency", weight = F)
```

The `amp_core` function also enables more refined core plots. Here each OTUs is plottted as "Abundant or not" vs. it's frequency. I've set the "Abundance cirteira" as 0.1 % abundance.
```{r core5, fig.width=6, fig.height=5, fig.align='center', message=FALSE}
amp_core(data=dsf, plot.type = "core", abund.treshold = 0.1)
```

All functions have an option to export the plot and the processed data. This is particulary usefull if you want to get the names and sequences of the "core" species. Here we store the results in the `res` object.
```{r core6}
res <- amp_core(data=dsf, plot.type = "core", abund.treshold = 0.1, output = "complete")
```

The processed data is stored as `data` in the output. Hence, to get the species that are abundant and seen in all samples we filter the data by `Frequency` and `freq_A`. Then we select the data we want to see, and finally sort it based on abundance.
```{r core7}
core_OTUs <- filter(res$data, Frequency > 51, freq_A > 51) %>% 
             select(OTU, Abundance, Phylum, Genus) %>%
             arrange(desc(Abundance))
```

```{r print_core}
print(core_OTUs)
```

As a last thing we want to extract the OTU sequences for further work. This can be done using the `amp_export` function. It simply takes a phyloseq object and exports the associated sequences. However, first we subset the phyloseq object to our "core OTUs" using the `prune_taxa` function from phyloseq.
```{r export1}
core_p <- prune_taxa(x = ds, taxa = as.character(core_OTUs$OTU))
```

.. Then the sequences are exported.
```{r export2}
amp_export(data = core_p, file = "my_sequences.fa")
```
